---
title: "ST558 Project 2"
author: "Crista Gregg, Halid Kopanski, Dionte Watie"
date: "7/2/2021"
params:
  day_of_week: "Monday"
output: 
  github_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(caret)
```
# Introduction

# Data
```{r message=FALSE}
set.seed(1)
bikes <- read_csv('day.csv')

day_function <- function(x){
  x <- as.character(x)
  switch(x, "0" = "Sunday", 
         "1" = "Monday", 
         "2" = "Tuesday", 
         "3" = "Wednesday", 
         "4" = "Thursday", 
         "5" = "Friday", 
         "6" = "Saturday")
}

bikes <- bikes %>% select(everything()) %>% mutate(weekday = sapply(weekday, day_function))

day <- params$day_of_week #will need to automate this

#filter bikes by day of week
bikes <- filter(bikes, weekday == day)
train_rows <- sample(nrow(bikes), 0.7*nrow(bikes))
train <- bikes[train_rows,]
test <- bikes[-train_rows,]
```

# Summarizations

## Contributions from Crista

The following table tells us the total number of rentals for each of the two years of collected data, as well as the average number of rentals per day.
```{r}
bikes %>%
  group_by(yr) %>%
  summarise(total_rentals = sum(cnt), avg_rentals = round(mean(cnt)))
```

The following box plot shows us how many rentals we have for days that are sunny or partly cloudy, misty, or rainy/snowy. We may expect some differences in behavior between weekend days where less people might be inclined to ride their bikes for pleasure, versus weekdays when more people might brave moderately unpleasant weather to get to work. 
```{r}
ggplot(bikes, aes(factor(weathersit), cnt)) +
  geom_boxplot() +
  labs(x = 'Type of Weather', y = 'Number of Rental Bikes', title = 'Rental Bikes by Type of Weather') +
  scale_x_discrete(labels = c('Clear to some clouds', 'Misty', 'Light snow or rain')) +
  theme_minimal()
```

Below is a chart of the relationship between casual and registered bikers. We might expect a change in the slope if we look at different days of the week. Perhaps we see more registered bikers riding on the weekday but more casual users on the weekend. 
```{r}
ggplot(bikes, aes(casual, registered)) +
  geom_point() +
  geom_smooth(method = 'lm') +
  theme_minimal() +
  labs(title = 'Registered versus Casual Renters')
```

Below we see a plot of the average daily number of bikers by month. We should expect to see more bikers in the spring and summer months, and the least in the winter. 
```{r}
plot_mth <- bikes %>%
  group_by(mnth) %>%
  summarize(avg_bikers = mean(cnt))

ggplot(plot_mth, aes(mnth, avg_bikers)) +
  geom_line(color = 'darkblue', size = 1.6) +
  theme_minimal() +
  labs(title='Average daily number of bikers by month', y = 'Average Daily Bikers', x = 'Month') +
  scale_x_discrete(limits = 1:12, labels = month.abb)
```

# Modeling

## Linear Regression

## Ensemble Tree

# Modeling

# Comparison